Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-02 doc/MS_v2.Rmd
updating manuscript

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    toc: yes
    toc_depth: '5'
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
always_allow_html: yes
bibliography: refs_Main_Report.bib
---

@@

*new* - The same dry-down spread technique was applied to both spermatocytes and oocytes based on (@peters_1997), with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used MLH1 quantification were between 5 and 12 weeks of age. Juvenile mice between 12 and 15 days of age were used for DMC1 quantification. Both ovaries were collected from embryos (16-21 embryonic days) or neonates (0-48 hours after birth). Whole testes were incubated in 3ml of hypotonic solution for 45 minutes. Decapsulated ovaries were incubated in 300ul of hypotonic solution for 45 minutes. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, cite). Slides were stored at -20 * C if not stained immediately.
To visualize the structure of meiotic chromosomes we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Crossovers (COs) were visualized as MLH1 foci. Double strand breaks (DSBs) were visualized as DMC1 foci. The staining protocol followed [@anderson1999] and [@koehler2002]. Antibody staining and slide blocking were performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmnuoResearch), 1X PBS, bovine serum albumin (Sigma), and Triton X-100 (Sigma)).
Following a 30-minute blocking wash in ABD, each slide was incubated with 60ul of a primary antibody master mix for 48 hours at 37* C. The master mix recipe contained polyclonal anti-rabbit anti-MLH1 (Calbiochem; diluted 1:50) or anti rabbit anti-DMC1) (mix of DMC1), anti-goat polyclonal anti-SYCP3, (Abcam; diluted 1:50), and anti-human polyclonal antibody to CREST (Antibodies, Inc; diluted 1:200) suspended in ADB. Slides were washed twice in 50ml ADB before the first round of secondary antibody incubation for 12 hours at 37* C. Alexa Fluor 488 donkey anti-rabbit IgG (Invitrgoen, location; diluted to 1:100) and Coumarin AMCA donkey anti-human IgG (Jackson ImmunoResearch; diluted to 1:200) were suspended in ADB. The last incubation of Alexa Fluor 568 donkey anti-goat (Invitrogen; diluted 1:100) was incubated at 1:100 for 2 hours at 37* C. Slides were fixed with Prolong Gold Antifade (Invitrogen) for 24 hours after a final wash in 1x PBS.

@@

*new-* To estimate the number of crossovers across the genome, we counted MLH1 foci. MLH1 foci were counted in cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. **Distributions of MLH1 count per cell were visually inspected for normality (supplemental figure )**. MLH1 foci located on the XY in spermatocytes were excluded from counts. In addition to MLH1 counts, we measured several traits to further characterize the recombination landscape. To estimate the number of double-strand breaks, a minority of which lead to crossovers, mean DMC1 foci per cell was quantified for a single male from a subset of strains ( _musculus^PWD^_, _musculus^MSM^_, _domesticus^WSB^_, and _domesticus^G^_ ). SC morphology and CREST foci number was used to stage spermatocytes as early zygotene or late zygotene.

@@

Total SC length per cell was quantified from pachtyene cell images [@wang2019_SC]. To reduce algorithmic errors in SC isolation, outliers were visually identified at the mouse level and removed from the data set. Mouse averages were calculated from cell-wide total SC lengths in `r formatC( length(  (Total.SC_DF  %>% filter(species == "M.musculus")  )$Random.Name), format="d", big.mark=",")` -- 3,371 out of `r formatC( length(  (MLH1_data %>% filter(species == "M.musculus") )$Random.Name), format="d", big.mark=",")`-- 4,143 cells with MLH1 counts.

@@

To examine crossover interference, the distance (in SC units) between MLH1 foci (inter-focal distance; IFD~raw~) was measured for those single bivalents containing two MLH1 foci. A normalized measure of interference (IFD~norm~) was computed by dividing IFDraw by SC length on a per-bivalent basis.

We used a series of statistical models to interpret patterns of variation in the recombination traits we measured (Table 2). We first constructed a mixed model (M1) using the lmer() from the lmer4 package (cite) in R (v3.5.2) [@Rstudio]. In this model, strain was coded as a random effect, with significance evaluated using a likelihood ratio test (using exactRLRT()). Subspecies, strain, and their interaction were coded as fixed effects,  with significance evaluated using a chi-square test comparing the full and reduced models (drop1() and anova()). *discribe the hiearchal nature*. As observations were at the level of single mouse (mouse average) was uniquely coded within the dataset, nesting was implict, each strain only occurs within one strain, and was not explictly coded in our linear and mixed models.

@@

```{r FOR.PLOT.organizing.data, echo=FALSE, message=FALSE, warning=FALSE}
#set the order of another column, based on another variable. consistant order for ploting

#apply this before ploting
MLH1_data <- MLH1_data %>%
  arrange(strain, sex, category, mouse) %>%
  mutate(Original.Name = factor(Original.Name)) #another category that you want the order to match
MLH1_data <- with(MLH1_data, MLH1_data[order(sex, strain),])

# sort your dataframe, by the focal categories
MLH1_data$mouse <- factor(MLH1_data$mouse, levels=unique(MLH1_data$mouse))
#use the above line for when order is out of wak

#MLH1_data <- with(MLH1_data, MLH1_data[order(category),])
Mouse.Table_HQ$var <- as.numeric(Mouse.Table_HQ$var)

AP_mouse_table.HQ_target <- Mouse.Table_HQ[!(is.na(Mouse.Table_HQ$sex) | Mouse.Table_HQ$sex==""), ]

remove_strains <- c("TOM","AST","CZECH","CAST","HMI","SPRET","CAROLI", "F1", "PERC")
AP_mouse_table.HQ_target <- AP_mouse_table.HQ_target[ ! AP_mouse_table.HQ_target$strain %in% remove_strains, ]

var.plot <- ggplot(AP_mouse_table.HQ_target, aes(y=var, x = sex, color=sex))+geom_jitter()+facet_wrap(~strain)+ggtitle("Plots of mouse Variances for MLH1 counts")+
  scale_color_manual(values=c("royalblue", "limegreen"))+theme_bw()

### Figure 1

```{r MLH1.MIXED.MODEL.setup, echo=FALSE, warning=FALSE, eval=TRUE, message=FALSE}

#DF.HetC.MixedModel.HQ
#this blck is duplicated in setupData.rm
#https://rcompanion.org/handbook/G_03.html
#testing random effects within lme

DF.HetC.MixedModel <- Mouse.Table_HQ[Mouse.Table_HQ$species == "M.musculus",]

#AP_mouse_table excluide Qual5 cells
DF.HetC.MixedModel.HQ <- Mouse.Table_HQ
DF.HetC.MixedModel.HQ <- add_species(DF.HetC.MixedModel.HQ)

#main df with variables coded as chr
DF.HetC.MixedModel.HQ$subsp <- factor(DF.HetC.MixedModel.HQ$subsp, ordered = FALSE)
DF.HetC.MixedModel.HQ$sex <- factor(DF.HetC.MixedModel.HQ$sex, ordered = FALSE)
DF.HetC.MixedModel.HQ$strain <- factor(DF.HetC.MixedModel.HQ$strain, ordered = FALSE)
#DF.HetC.MixedModel.HQ$species <- factor(DF.HetC.MixedModel.HQ$species, ordered = FALSE)

#kable(table(DF.HetC.MixedModel.HQ$sex, DF.HetC.MixedModel.HQ$strain))

all.cell.MM.HQ <- ddply(.data=DF.HetC.MixedModel.HQ,
                 .(sex, strain),
                 summarize,
                 all.cells = sum(Ncells)
)

@@

```{r Fig1.top, echo=FALSE, warning=FALSE, message=FALSE}
#Mouse.Table_HQ
#change the segements and text!!

Fig1_v1_strain <- ggplot(strain.table.HQ_Fig, aes(y=mean_co, x=strain, color=strain, shape=sex))+ 
  ylab("Average MLH1 Foci per Cell")+

  geom_point(aes(shape = sex), size=3, position=position_dodge(.5))+

    geom_errorbar(aes(ymin=mean_co-(se*2), ymax=mean_co+(se*2) ), width=.3, position = position_dodge(.5), size=1)+

  ylim(c(18, 35))+
 scale_color_manual(values=colors_of_strains)+

    theme(legend.position="none",
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"),
  axis.title.y=element_blank(),
  axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank())+
#remove y labfor male

   #mol
  annotate("segment", x=.75, xend=2.3, y=19.5, yend=19.5, colour="black", size=1)+

  #Musc,
  annotate("segment", x= 2.75, xend=5.2, y=30, yend=30, colour="black", size=1)+

  #dom
  annotate("segment", x=5.8, xend=8.5, y=19.5, yend=19.5, colour="black", size=1)+

  #cast
  annotate("segment", x=8.8, xend= 9.3, y=30, yend=30, colour="black", size=1)+

    #spret
  annotate("segment",x=9.5, xend=10.2, y=19.5, yend=19.5, colour="black", size=1)+

  #spic
  annotate("segment",x=10.5, xend=11.2, y=30, yend=30, colour="black", size=1)+

  #label
    annotate("text", x = c(1.6,3.9,6.3,8,10, 10.3), #list of all x points 
           y = c(18.5, 31, 18.5, 31, 18.5, 31),

   label = c("M. m. molossinus","M. m. musculus", "M. m. domesticus", "M. m. castaneus", "M. spretus", "M. spicilegus"),
           fontface = 'italic', hjust = 0.5, size=3.5)

```{r female.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5, warning=FALSE, message=FALSE}

#MLH1_data  %>% filter(quality < 5)
#mouse avs

female_mouse.av <- Mouse.Table_HQ.OG %>% filter(sex == "female") %>% filter(species == "M.musculus")

female_mouse.av <- add_sex(female_mouse.av)
female_mouse.av <- add_strain(female_mouse.av)
female_mouse.av <- add_subsp(female_mouse.av)
female_mouse.av <- add_species(female_mouse.av)

female_mouse.av$strain <- factor(female_mouse.av$strain, ordered = FALSE, levels =c( "MSM", "MOLF",
  "PWD", "SKIVE", "KAZ",

  "WSB", "G", "LEW",
                                                                         "CAST", 
                                                                   "SPRET", "SPIC") )

female.specific.plot <- ggplot(female_mouse.av, aes(x = as.factor(strain), y = mean_co, color= strain))+ geom_boxplot()+  geom_jitter(aes( alpha=0.5), width = .2 )+ ggtitle("")+scale_color_manual(values=colors_of_strains)+ylim(c(17,34))+
  theme(legend.position="none",

        axis.title.y =element_blank(), 
      axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"))+geom_hline(yintercept = 20)

#+theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r male.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5, warning=FALSE, message=FALSE}
#try customize the plot size

male_mouse.avs_for.analysis$strain <- factor(male_mouse.avs_for.analysis$strain, ordered = FALSE, levels =c( "MSM", "MOLF",
  "PWD", "SKIVE", "KAZ","TOM","AST","CZECH",

  "WSB", "G", "LEW","PERC",
                                                                         "CAST", "HMI",
                                                                   "SPRET", "SPIC") )

male.plot <- ggplot( male_mouse.avs_for.analysis, aes(x = as.factor(strain), y = mean_co, color= strain))+geom_jitter( alpha=0.5, width = .2) +geom_boxplot()+ ggtitle("")+scale_color_manual(values=colors_of_strains)+ylim(c(17,34))+

   theme(legend.position="none", axis.text.x=element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),  axis.title.y = element_blank(), 
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"))+geom_hline(yintercept = 19)

#horizontal line = 20

```{r MLH1.cell.examples, echo=FALSE, message=FALSE, warning=FALSE}

male.MLH1.cell.ex <- readPNG('~/MLH1repo/doc/figureFiles/MLH1.male.cell_ex.png')

female.MLH1.cell.ex <- readPNG('~/MLH1repo/doc/figureFiles/MLH1.female_cell.ex.png')

empty.DF <- data.frame(col1 = c(1,3), col2 = c(1,3) )

#Cell GGPLOT

maleMLH1.cell.obj <- ggplot(empty.DF, aes(x = col1, y = col2)) +
  background_image(male.MLH1.cell.ex) + theme(axis.title.x=element_blank(),
     axis.text.x=element_blank(),axis.ticks.x=element_blank(),axis.title.y=element_blank(),                                  axis.text.y=element_blank(),axis.ticks.y=element_blank() )

femaleMLH1.cell.obj <- ggplot(empty.DF, aes(x = col1, y = col2)) +
  background_image(female.MLH1.cell.ex) + theme(axis.title.x=element_blank(),
     axis.text.x=element_blank(),axis.ticks.x=element_blank(),axis.title.y=element_blank(),                                  axis.text.y=element_blank(),axis.ticks.y=element_blank() )

```{r Figure1.Main.plot.three.plots, echo=FALSE, warning=FALSE, fig.width=7, fig.cap='*Figure 1 Mean MLH1 count distributions by strain.* A) Strain averages of MLH1 counts per cell. B) Female specific MLH1 count distributions. C)	Male specific MLH1 counts per cell.', message=FALSE}

MLH1.Figure1  <- ggdraw() +
  draw_plot(Fig1_v1_strain, x = .02, y = .55, width=1, height=.45)+

  draw_plot(female.specific.plot, x=0, y=0, width=.5,height= .55)+
  draw_plot(femaleMLH1.cell.obj, x=.35, y=.35, width=.18, height= .18)+

  draw_plot(male.plot, x=.5, y = 0,  width = .5, height = .55)+
  draw_plot(maleMLH1.cell.obj, x=.85, y = .35,  width = .19, height=.19)+

  draw_plot_label(label = c("A", "B", "C"), size = 14,
                  x = c(0, .01, .51), y = c(.99, .55, 0.55) )

MLH1.Figure1
#add the cell images

#ggsave("~/MLH1repo/doc/Figure1.png")
```

## B Genome-wide recombination rate evolves differently in females and males

@@

```{r female.strain.table, echo=FALSE}
female_mouse.avs_for.analysis <- female_mouse.avs_for.analysis %>% filter(strain != "CAST")
#female strain table
#female_mouse.avs_for.analysis
F.strain.table <- ddply(female_mouse.avs_for.analysis, c("strain"), summarise,

                        Nmice = length(unique(mouse)),

                        total.cells  = sum(Ncells, na.rm = TRUE),

                        mouse.mean_co = mean(mean_co, na.rm=TRUE),

                        mean.mouse_cV = mean(cV, na.rm = TRUE),

                        mean.mouse_var = mean(var, na.rm = TRUE ),

                        mean.mouse_sd   = mean(sd, na.rm = TRUE ),

                        #Standard error (mice vs cells)
                        mean.mouse_se  = mean(se, na.rm=TRUE)
    )
#F.strain.table

#F.strain.table

# I don't know why most of the table is blank
```

#### Mouse variance plots

```{r mouse.var.plots.female, echo=FALSE, fig.width = 4, fig.height=2.5, include=FALSE, eval=FALSE}

#variance plots
musc.female.VAR.plot <- ggplot( (female_mouse.avs_for.analysis %>% filter(sex== "female") %>% filter(subsp == "Musc") ) , aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Musc female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
musc.female.VAR.plot

Mol.female.VAR.plot <- ggplot( (female_mouse.avs_for.analysis %>% filter(subsp == "Mol") ), aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Mol female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Mol.female.VAR.plot

Dom.female.VAR.plot <- ggplot(Dom.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain), width = .2 )+ ggtitle("Dom female\nwithin mouse  MLH1 distributions") + scale_color_manual(values=colors_of_strains)
Dom.female.VAR.plot

```{r male.strain.table, echo=FALSE}

M.strain.table <- ddply(male_mouse.avs_for.analysis, c("strain"), summarise,
                        Nmice =length(unique(mouse)),

                        totNcells  = sum(Ncells, na.rm=TRUE),

                        mean.mouse_mean_co = mean(mean_co, na.rm=TRUE),

                        mean.mouse_cV = mean(cV, na.rm = TRUE),

                        mean.mouse_var = mean(var, na.rm = TRUE ),

                        mean.mouse_sd   = mean(sd, na.rm=TRUE),

                        #Standard error (mice . cells)
                        mean.mouse_se  = mean(se, na.rm=TRUE)
    )
#M.strain.table
#same issue as the female table .. ugh
```

```{r mouse.var.plots.male, echo=FALSE, fig.width = 4, fig.height=2.5, include=FALSE, include=FALSE, eval=FALSE}

#variance plots
musc.male.VAR.plot <- ggplot(Musc.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Musc male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)

Mol.male.VAR.plot <- ggplot(Mol.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Mol male of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)

Dom.male.VAR.plot <- ggplot(Dom.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("VAR Dom male ofwithin mouse  MLH1 distributions")+scale_color_manual(values=colors_of_strains)

@@

Main table is strain level, (n)

@@

To further situate variation in house mouse recombination rate within an evolutionary framework, we fit a series of statistical models including subspecies, strain, and sex, to mean MLH1 focus counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` -- 187 mice. **to best investigate sex, will limited our analysis to strains with sufficient obersevations from both sexes (Table 1).**

@@

These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains. Analysis of the female dataset identifies four strains with effects on recombination rate: _domesticus^G^_ (p = 2.510^{-6} -- `r sum_MLH1.female.spc.glm.M2$coefficients[2,][4]`),  *why is mol so different*  _molossinus^MSM^_  (p = 6.2410^{-6} -- `r sum_MLH1.female.spc.glm.M2$coefficients[5,][4]`), _domesticus^LEW^_ (p = 0.01 -- `r sum_MLH1.female.spc.glm.M2$coefficients[3,][4]` ), and _musculus^PWD^_ (p= 0.02 -- `r sum_MLH1.female.spc.glm.M2$coefficients[4,][4]` ). Strain effect sizes in females are modest in magnitude (ranging from 1 to 4 foci) compared to those in males. 
Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.

## B Synaptonemal complexes are longer in females

*new* -- Our survey of recombination rate across house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (all t-test below; p < 0.05; except short bivalents for SKIVE p = `r Q1_short.SC_t.test.SKIVE$p.value` ).

In the dataset of short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from 1.15 -- `r Q1SC.mus.av.short.biv_DIFF_MSM` ( _musculus^MSM^_) to 1.49 -- `r Q1SC.mus.av.short.biv_DIFF_WSB` ( _domesticus^WSB^_) across strains.

That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC  (M1: p=`r Q1.lmer_totSC_M1_results$sex.results`; M2: p=`r sum.Q1.glm_totSC_M2$coefficients[,c(1,4)][8,2]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,c(1,4)][8,2]`) and short bivalent measures (M1: p= `r Q1.lmer_short_M1_results$sex.results`; M2: p= `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][8,]`; M3: `r summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)][8,2]`).

@@

## B Females and males differ in crossover positions and crossover interference

### triangles

double triangle plots made with different scripts

```{r double.IFD.triangle.plots, echo=FALSE, eval=FALSE}

#good function, saves to file, can't reproduce in rmd ..width=650, height=500
#big loop for automating the IFD triangles
only.male.strains <- c("F1", "PERC", "CZECH", "AST", "TOM", "HMI", "SPRET", "CAROLI", "other")
auto_curate_DF <- Curated_BivData[ ! Curated_BivData$strain %in% only.male.strains, ]

df_strain_list <- split(auto_curate_DF, (auto_curate_DF$strain))
df_strain_list <- df_strain_list[ -c(4,10,11,12,13,14,15,16,17,18,19) ]
#make a new empty df. create a ddply thing with mean calculations (this is like a list for matching up later)

#this works
top.triangle <- function(my.y, my.x, my.data) {
  plot(my.y ~ my.x, data= my.data, axes=FALSE, ylab='', xlab="", 
       main='', xlim=c(0, 1), ylim=c(0, 1), xaxt="n", yaxt="n", asp=1,
       col =  colors_of_strains[current.strain])

  mtext(current.strain, 3, 5, font=2, cex=1.3, adj=.95)
  mtext("Position.2.Female", 2, .75)
  mtext("Position.1.Female", 3, 2)

  axis(side=2, las=1, pos=0)
  axis(side=3, las=1, pos=1)
  lines(0:1, 0:1)
}

bottom.triangle <- function(my.y, my.x, my.data){
  points(my.x ~ my.y, data=my.data, xpd=TRUE, 
          col =  colors_of_strains[current.strain])

  mtext("Position.2.Male", 1, 1.5, at=mean(par()$usr[1:2]) + x.dist)
  mtext("Position.1.Male", 4, 3, padj=par()$usr[1] +  7)
  x.at <- axisTicks(par()$usr[1:2], 0) + x.dist
  axis(side=1, las=1, pos=0, at=x.at, 
       labels=F, xpd=TRUE)
  mtext(seq(0, 1, .2), 1, 0, at=x.at)
  axis(4, las=1, pos=1 + x.dist)
  lines(0:1 + x.dist, 0:1, xpd=TRUE)
}

x.dist <- .1
RP <- list()

for(i in 1:(length(df_strain_list))){
  #print(i)
  current.strain <- as.character(df_strain_list[[i]]$strain[1])

  #declare file for saving png
  #png(paste0("~./MLH1repo/doc/figureFiles/IFD.triangle_", current.strain, ".png"),  width=650, height=500, res=100)

  plot.new()

  my.data = (Curated_BivData %>% filter(strain == current.strain) %>% filter(hand.foci.count == 2) )

  my.x.top = (Curated_BivData %>% filter(strain == current.strain) %>% filter(sex == "female") %>% filter(hand.foci.count == 2) )$PER_Foci_1

  my.y.top = (Curated_BivData %>% filter(strain == current.strain) %>% filter(sex == "female") %>% filter(hand.foci.count == 2) )$PER_Foci_2

  my.x.bottom = (Curated_BivData %>% filter(strain == current.strain) %>% filter(sex == "male") %>% filter(hand.foci.count == 2) )$PER_Foci_1

  my.y.bottom = (Curated_BivData %>% filter(strain == current.strain) %>% filter(sex == "male") %>% filter(hand.foci.count == 2) )$PER_Foci_2

  #adjusting this mar lowers the margins
  op <- par(mar=c(.6,.6, .6, .6) + 0.1, oma=c(2, 0, 0, 2))

  top.triangle(my.y.top, my.x.top, my.data)

  bottom.triangle(my.y.bottom+x.dist, my.x.bottom, my.data) #x.dist

  par(op)
  RP[[i]] <- recordPlot()
  dev.off()
}
```

```{r Q1.IFD_load.triangles, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "Distribution of normalized foci positions (IFDs) for double crossover bivalents across sexes. Female and male patterns are displayed in the top and bottom triangle respectively. Each point reflects a single double crossover bivalent, with the normalized positions of foci reflected on each axis."}

plot1.WSB <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_WSB.png')
plot2.LEW <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_LEW.png')
plot3.G <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_G.png')
plot4.KAZ <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_KAZ.png')
plot5.SKIVE <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_SKIVE.png')
plot6.PWD <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_PWD.png')
plot7.MSM <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_MSM.png')

# grid.arrange(rasterGrob(plot1.WSB), rasterGrob(plot2.LEW), rasterGrob(plot3.G),
#             rasterGrob(plot4.KAZ), rasterGrob(plot5.SKIVE), rasterGrob(plot6.PWD),
#             rasterGrob(plot7.MSM), 
#             ncol=3)

#load all the files / arrange the plots from above 
#WSB.triangle.png <- readPNG("~./MLH1repo/doc/figureFiles/test.triangle_WSB.png")
```

@@

## B Evolution of genome-wide recombination rate is dispersed across bivalents, associated with double-strand break number, and connected to crossover interference

turn off all these DMC1 tables -- to makes sure the right DFs are loaded in from the setup scripts.

### Figure 2 DMC1

>>>>

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  word_document:
  pdf_document:
    toc: yes
    toc_depth: '5'
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
always_allow_html: yes
bibliography: refs_Main_Report.bib
---

@@

The same dry-down spread technique was applied to both spermatocytes and oocytes based on (@peters_1997), with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used MLH1 quantification were between 5 and 12 weeks of age. Juvenile mice between 12 and 15 days of age were used for DMC1 quantification. Both ovaries were collected from embryos (16-21 embryonic days) or neonates (0-48 hours after birth). Whole testes were incubated in 3ml of hypotonic solution for 45 minutes. Decapsulated ovaries were incubated in 300ul of hypotonic solution for 45 minutes. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, cite). Slides were stored at -20 * C if not stained immediately.
To visualize the structure of meiotic chromosomes we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Crossovers (COs) were visualized as MLH1 foci. Double strand breaks (DSBs) were visualized as DMC1 foci. The staining protocol followed [@anderson1999] and [@koehler2002]. Antibody staining and slide blocking were performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmunoResearch), 1X PBS, bovine serum albumin (Sigma), and Triton X-100 (Sigma)).
Following a 30-minute blocking wash in ABD, each slide was incubated with 60ul of a primary antibody master mix for 48 hours at 37* C. The master mix recipe contained polyclonal anti-rabbit anti-MLH1 (Calbiochem; diluted 1:50) or anti rabbit anti-DMC1) (mix of DMC1), anti-goat polyclonal anti-SYCP3, (Abcam; diluted 1:50), and anti-human polyclonal antibody to CREST (Antibodies, Inc; diluted 1:200) suspended in ADB. Slides were washed twice in 50ml ADB before the first round of secondary antibody incubation for 12 hours at 37* C. Alexa Fluor 488 donkey anti-rabbit IgG (Invitrgoen, location; diluted to 1:100) and Coumarin AMCA donkey anti-human IgG (Jackson ImmunoResearch; diluted to 1:200) were suspended in ADB. The last incubation of Alexa Fluor 568 donkey anti-goat (Invitrogen; diluted 1:100) was incubated at 1:100 for 2 hours at 37* C. Slides were fixed with Prolong Gold Antifade (Invitrogen) for 24 hours after a final wash in 1x PBS.

@@

*new-* To estimate the number of crossovers across the genome, we counted MLH1 foci. MLH1 foci were counted in cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. Distributions of MLH1 count per cell were visually inspected for normality (Supplemental figure 1). MLH1 foci located on the XY in spermatocytes were excluded from counts. In addition to MLH1 counts, we measured several traits to further characterize the recombination landscape. To estimate the number of double-strand breaks, a minority of which lead to crossovers, mean DMC1 foci per cell was quantified for a single male from a subset of strains ( _musculus^PWD^_, _musculus^MSM^_, _domesticus^WSB^_, and _domesticus^G^_ ). SC morphology and CREST foci number was used to stage spermatocytes as early zygotene or late zygotene.

@@

Total SC length per cell was quantified from pachytene cell images [@wang2019_SC]. To reduce algorithmic errors in SC isolation, outliers were visually identified at the mouse level and removed from the data set. Mouse averages were calculated from cell-wide total SC lengths in `r formatC( length(  (Total.SC_DF  %>% filter(species == "M.musculus")  )$Random.Name), format="d", big.mark=",")` -- 3,371 out of `r formatC( length(  (MLH1_data %>% filter(species == "M.musculus") )$Random.Name), format="d", big.mark=",")`-- 4,143 cells with MLH1 counts.

@@

To examine crossover interference, the distance (in SC units) between MLH1 foci (inter-focal distance; IFD~raw~) was measured for those single bivalents containing two MLH1 foci. A normalized measure of interference (IFD~norm~) was computed by dividing IFD~raw~ by SC length on a per-bivalent basis.

We used a series of statistical models to interpret patterns of variation in the recombination traits we measured (Table 2). We first constructed a mixed model (M1) using the lmer() from the lmer4 package (cite) in R (v3.5.2) [@Rstudio]. In this model, strain was coded as a random effect, with significance evaluated using a likelihood ratio test (using exactRLRT()). Subspecies, strain, and their interaction were coded as fixed effects,  with significance evaluated using a chi-square test comparing the full and reduced models (drop1() and anova()). As observations were at the level of single mouse (mouse average) was uniquely coded within the dataset, nesting was implicit, each strain only occurs within one strain, and was not explicitly coded in our linear and mixed models.

@@

@@

@@

@@

## Genome-wide recombination rate evolves differently in females and males

@@

To further situate variation in house mouse recombination rate within an evolutionary framework, we fit a series of statistical models including subspecies, strain, and sex, to mean MLH1 focus counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` -- 187 mice. **to best investigate sex, will limited our analysis to strains with sufficient observations from both sexes (Table 1).**

@@

These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains. Analysis of the female dataset identifies four strains with effects on recombination rate: _domesticus^G^_ (p = 2.510^{-6} -- `r sum_MLH1.female.spc.glm.M2$coefficients[2,][4]`), _molossinus^MSM^_  (p = 6.2410^{-6} -- `r sum_MLH1.female.spc.glm.M2$coefficients[5,][4]`), _domesticus^LEW^_ (p = 0.01 -- `r sum_MLH1.female.spc.glm.M2$coefficients[3,][4]` ), and _musculus^PWD^_ (p= 0.02 -- `r sum_MLH1.female.spc.glm.M2$coefficients[4,][4]` ). Strain effect sizes in females are modest in magnitude (ranging from 1 to 4 foci) compared to those in males. 
Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.

## Synaptonemal complexes are longer in females

Our survey of recombination rate across house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (all t-test below; p < 0.05; except short bivalents for SKIVE p = `r Q1_short.SC_t.test.SKIVE$p.value` ).

In the dataset of short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from 1.15 -- `r Q1SC.mus.av.short.biv_DIFF_PWD` ( _musculus^PWD^_) to 1.49 -- `r Q1SC.mus.av.short.biv_DIFF_WSB` ( _domesticus^WSB^_) across strains.

That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC  (M1: p=`r Q1.lmer_totSC_M1_results$sex.results`; M2: p=`r sum.Q1.glm_totSC_M2$coefficients[,c(1,4)][9,2]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,c(1,4)][8,2]`) and short bivalent measures (M1: p= `r Q1.lmer_short_M1_results$sex.results`; M2: p= `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][8,2]`; M3: `r summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)][8,2]`).

@@

## Females and males differ in crossover positions and crossover interference

@@

## Evolution of genome-wide recombination rate is dispersed across bivalents, associated with double-strand break number, and connected to crossover interference
